
# Phylogenetically independent contrasts

Loading required packages:

```{r load-pkgs}
#| message: false
#| warning: false

set.seed(123) # for reproducibility

library(phytools)
library(ape)
library(ggplot2)
library(patchwork)
```

## Practice problems

1. Repeat your PIC analysis comparing body mass and home range but for 
only one clade in your tree: the artiodactyls. Use `extract.clade()` 
to pull out this taxon from your tree of mammals. Do the results 
and conclusions of your analysis change? If so, in what way?

```{r fig.height=9, fig.width=8}
#| fig-width: 8
#| fig-height: 9

# Loading data
mammalHR <- read.csv(
    "http://www.phytools.org/Rbook/2/mammalHR.csv", row.names = 1
)
mammal_tree <- read.tree("http://www.phytools.org/Rbook/2/mammalHR.phy")

# Visualize the tree to see the node number for Artiodactyla
plotTree(mammal_tree, ftype = "i")
nodelabels(bg = "white", cex = 0.5, frame = "circle")

# Extract Artiodactyla (node 75)
artio_tree <- extract.clade(mammal_tree, node = 75)
artio_tree

# Subset the phenotypic data to include only Artiodactyla species
artio <- mammalHR[artio_tree$tip.label, ]
head(artio)

# Get home range and bodty mass as named vectors
home_range <- setNames(artio$homeRange, rownames(artio))
home_range

body_mass <- setNames(artio$bodyMass, rownames(artio))
body_mass
```

```{r}
# Compute PICs for home range and body mass
pic_homerange <- pic(log(home_range), artio_tree)
pic_bodymass <- pic(log(body_mass), artio_tree)

pic_homerange
pic_bodymass


# Fit linear model to PICs without intercept
pic_model <- lm(pic_homerange ~ pic_bodymass + 0)
summary(pic_model)

# Visualizing the model
p_pics <- data.frame(
    Home_range = pic_homerange, 
    Body_mass = pic_bodymass
) |>
    ggplot(aes(x = Body_mass, y = Home_range)) +
    geom_point(size = 3, color = "deepskyblue4") +
    geom_vline(xintercept = 0, linetype = 3) +
    geom_hline(yintercept = 0, linetype = 3) +
    geom_smooth(method = "lm", se = FALSE, formula = "y ~ x") +
    labs(
        title = "Regression of home range on body size for Artiodactyla",
        subtitle = "Regression was performed on phylogenetically independent contrasts (PICs)",
        x = "PICs for log(body mass)",
        y = "PICs for log(range size)",
    ) +
    theme_classic()

p_pics
```

Results and conclusions did not change.


2. On the book website, we supply some data for a group of Asian barbets (`BarbetTree.nex` and `Barbetdata_mod.csv`): members of the bird family 
Megalaimidae [@gonzalez2013evolution]. Using these data and the method of
phylogenetic independent contrasts, test for a relationship between 
the two variables *Lnalt* and *wing*. 
What do you find? How does the slope of your relationship change 
if you fit *Lnalt wing* versus *wing Lnalt*? 
How does the P-value of a statistical test of this relationship change?

```{r fig.width=8, fig.height=5}
# Get data
barbet <- read.csv(
    "http://www.phytools.org/Rbook/2/Barbetdata.csv", 
    row.names = 1
)
barbet_tree <- read.nexus("http://www.phytools.org/Rbook/2/BarbetTree.nex")

# Take a look at the data
head(barbet)
barbet_tree

# Check tree and data for inconsistencies
check <- geiger::name.check(barbet_tree, barbet)
check

barbet_tree <- drop.tip(barbet_tree, check$tree_not_data)

# Define function to fit regression on PICs for variables x and y
fit_pics <- function(tree, data, x, y) {
    
    # Get variables as named vectors
    var_x <- log(setNames(data[[x]], rownames(data)))
    var_y <- log(setNames(data[[y]], rownames(data)))

    # Compute PICs
    pic_x <- pic(var_x, tree)
    pic_y <- pic(var_y, tree)
    
    # Fit linear model to PICs without intercept
    pic_model <- summary(lm(pic_y ~ pic_x + 0))
    
    # Get estimates
    r2 <- signif(pic_model$r.squared, 1)
    slope <- signif(pic_model$coefficients[1], 2)
    p <- format(pic_model$coefficients[4], scientific = TRUE)
    
    subtitle <- bquote(beta[1] == .(slope) * ", " ~ R^2 == .(r2) * ", " ~ P == .(p))

    # Visualizing the model
    p_pics <- data.frame(
        X = pic_x, 
        Y = pic_y
    ) |>
        ggplot(aes(x = X, y = Y)) +
        geom_point(size = 3, color = "deepskyblue4") +
        geom_vline(xintercept = 0, linetype = 3) +
        geom_hline(yintercept = 0, linetype = 3) +
        geom_smooth(method = "lm", se = FALSE, formula = "y ~ x") +
        labs(
            title = paste0("Regression of ", y, " on ", x),
            subtitle = subtitle,
            x = paste0("PICs for log(", x, ")"),
            y = paste0("PICs for log(", y, ")")
        ) +
        theme_classic()
    
    p_pics
    return(p_pics)
}

# Fit a regression of `wing` on `Lnalt` using PICs
model1 <- fit_pics(barbet_tree, barbet, "Lnalt", "wing")

# Fit a regression of `Lnalt` on `wing` using PICs
model2 <- fit_pics(barbet_tree, barbet, "wing", "Lnalt")

# Visualize plots together
wrap_plots(
    model1, model2, ncol = 2
)
```

In both models (regressing `wing` on `Lnalt` and regressing `Lnalt` on
`wing`), there is no significant association association between
the variables. The P-value and variance explained remain the same,
but the slope ($B_1$) changes, which is expected considering that predictors
and outcomes have been flipped.


3. If you multiply all the branches of your phylogenetic tree by 100, 
will your independent contrasts analysis change? Why or why not? 
Can you confirm this using R code? Don’t forget to use what you learned
about the internal structure of a "phylo" object in chapter 1 to 
modify your tree.

```{r}
# Multiple all branches by 100
new_barbet_tree <- barbet_tree
new_barbet_tree$edge.length <- new_barbet_tree$edge.length * 100

# Compute PICs for `wing` using original and modified tree
original_pic_wing <- pic(
    log(setNames(barbet$wing, rownames(barbet))), barbet_tree
)
modified_pic_wing <- pic(
    log(setNames(barbet$wing, rownames(barbet))), new_barbet_tree
)

# Are PICs the same?
identical(original_pic_wing, modified_pic_wing)
head(original_pic_wing)
head(modified_pic_wing)

# Perform regression of `wing` on `Lnalt` using both trees
original_model <- fit_pics(barbet_tree, barbet, "Lnalt", "wing")
modified_model <- fit_pics(new_barbet_tree, barbet, "Lnalt", "wing")

# Compare models
wrap_plots(original_model, modified_model)
```

The PICs change (PICs using the modified tree are the PICs of the original
tree divided by 10), but the results of the regression using PICs do not 
change, because the relationship between the species (i.e., the tree topology)
remained the same after multiplying branch lengths by 100.


## References {.unnumbered}
