
# A brief introduction to phylogenetics in R

Loading required packages:

```{r load-pkgs}
#| message: false
#| warning: false

# Load required packages
library(ape)
library(geiger)
library(phytools)
library(ggplot2)
library(patchwork)
```

## Practice problems

1. Download the following two files for *Phelsuma geckos* from the book
website: `phel.csv` and `phel.phy` (Harmon et al. 2010). 
`phel.csv` is a CSV file containing trait values for ten different 
morphological traits. `phel.phy` is a phylogeny of thirty-three species. 
Read both data and tree in from file and use `name.check()` to identify 
any differences between the two data sets. If you find differences, prune the
phylogeny and subsample the trait data to include only the species present in 
both the data file and the tree. Plot the tree.

```{r p1, fig.height=6, fig.width=6}
#| fig-height: 6
#| fig-width: 6

# Get data
pheno_data <- read.csv(
    "http://www.phytools.org/Rbook/1/phel.csv", row.names = 1
)
tree <- read.tree("http://www.phytools.org/Rbook/1/phel.phy")

# Taking a look at the data
head(pheno_data)
tree

# Checking for differences between the data sets
check <- name.check(tree, pheno_data)
check

# Pruning the tree to match data
final_tree <- drop.tip(tree, check$tree_not_data)
final_tree

# Keep in the data frame only taxa that are in the tree
final_pheno <- pheno_data[final_tree$tip.label, ]
head(final_pheno)

# Plot the tree
plot(final_tree)
```

2. Use `phyl.pca()` to run a phylogenetic principal component analysis (PCA) 
of the morphological data set and tree from practice problem 1. When data for
different variables in a PCA have different orders of magnitude, 
it often makes sense to transform by the natural logarithm and 
conduct our analysis on the log-transformed values instead of on the original 
traits. Inspect your data to see if this applies and then decide whether or 
not to log-transform before undertaking your phylogenetic PCA. After you’ve 
obtained a result for the PCA, create a screeplot to visualize the 
distribution of variation between the different principal component axes.

```{r p2}
# Perform phylogenetic PCA
pca <- phyl.pca(final_tree, final_pheno)
pca

# Create a screeplot with variance explained by each PC
p_screeplot <- data.frame(
    Variance = signif(as.princomp(pca)$sdev ^ 2, 2),
    PC = factor(paste0("PC", 1:10), levels = paste0("PC", 1:10))
) |>
    ggplot(aes(x = PC, y = Variance * 100)) +
    geom_col(fill = "dodgerblue4", color = "black") +
    geom_text(aes(label = Variance), vjust = -0.3) +
    ylim(0, 100) +
    labs(
        title = "Screeplot for phylogenetic PCA of Phelsuma sp. morphological traits",
        x = "", y = "Variance explained (%)"
    ) +
    theme_bw()

p_screeplot
```

3. Use `phylomorphospace()` to create a single projection of the phylogeny 
into morphospace for the first two PC axes from practice problem 1.2. Can you 
think of a way to project the tree into a space defined by more than two 
principal component dimensions? Hint: look up the help pages for 
`phylomorphospace3d()` and `phyloScattergram()` for ideas, or consider 
simply subdividing your plotting device using `par(mfrow)`.

```{r}
# Create phylomorphospace plot for different pairwise combinations of PCs
combs <- list(
    c(1, 2),
    c(1, 3),
    c(2, 3),
    c(1, 4)
)

p_pms <- lapply(combs, function(x) {
    ## Get phylomorphospace coordinates
    score_df <- as.data.frame(scores(pca)[, x])
    cols <- names(score_df)
    
    pms <- phylomorphospace(
        final_tree, score_df
    )
    coord <- data.frame(
        xstart = pms$xx[pms$edge[, 1]],
        ystart = pms$yy[pms$edge[, 1]],
        xstop = pms$xx[pms$edge[, 2]],
        ystop = pms$yy[pms$edge[, 2]],
        nodestart = pms$edge[, 1],
        nodestop = pms$edge[, 2]
    )
    
    p_phylomorphospace <- ggplot() +
        geom_point(
            data = score_df, 
            aes(x = .data[[cols[1]]], y = .data[[cols[2]]]), 
            size = 3, color = "darkblue"
        ) +
        geom_segment(
            data = coord, aes(x = xstart, y = ystart, xend = xstop, yend = ystop), 
            size = 0.5
        ) +
        theme_classic() +
        labs(x = cols[1], y = cols[2])
    
    return(p_phylomorphospace)
})

# Arrange plots in a panel
wrap_plots(p_pms) +
    plot_annotation(
        title = "Phylomorphospace of different pairwise combinations of PCs"
    )
```


## References {.unnumbered}