
# Phylogenetic generalized least squares

Loading required packages:

```{r load-pkgs}
#| message: false
#| warning: false

set.seed(123) # for reproducibility

library(phytools)
library(ape)
library(nlme)
library(ggplot2)
library(patchwork)
library(dplyr)
```

## Practice problems

1. In practice problem 2 of the previous chapter, you used Asian barbet data 
to test whether the variable *Lnalt* varied as a function of *wing*. 
Again, using these data from Asian barbets (`BarbetTree.nex` 
and `Barbetdata_mod.csv`), carry out the same analysis using PGLS. 
Confirm that you get the same results from PGLS and PICs. 
What happens to your fitted model if you also estimate $\lambda$ using 
the `corPagel()` function?

```{r}
#| warning: false

# Load data
barbet <- read.csv(
    "http://www.phytools.org/Rbook/2/Barbetdata.csv", 
    row.names = 1
)
barbet_tree <- read.nexus("http://www.phytools.org/Rbook/2/BarbetTree.nex")

# Fix inconsistencies
check <- geiger::name.check(barbet_tree, barbet)
barbet_tree <- drop.tip(barbet_tree, check$tree_not_data)

# Define a function to fit PGLS
fit_pgls <- function(
        tree, data, 
        model_formula = NULL, 
        corStruct = NULL, 
        gls_method = "ML"
    ) {
    
    # Fit PGLS
    model <- nlme::gls(
        as.formula(model_formula), data = data, 
        correlation = corStruct, 
        method = gls_method
    )
    output <- summary(model)
    
    return(output)
}

# Fit regression on PICs with PGLS
source(here::here("utils.R"))

pic_model <- fit_pics(
    tree = barbet_tree, 
    data = barbet, 
    x = "wing", y = "Lnalt", 
    return = "model"
)

spp <- rownames(barbet)
pgls_model <- fit_pgls(
    tree = barbet_tree, 
    data = log(barbet), 
    model_formula = "Lnalt ~ wing", 
    corStruct = corBrownian(phy = barbet_tree, form = ~spp)
)

# Compare models
model_comparison <- data.frame(
    Model = c("PIC", "PGLS"),
    Slopes = c(pic_model$coefficients[1], pgls_model$coefficients[2]),
    P_values = c(pic_model$coefficients[4], pgls_model$tTable[2, 4]),
    row.names = NULL
)
model_comparison
```

As we can see, the slopes ($\beta_1$) and P-values are numerically identical,
which confirms that regression on PICs and PGLS are equivalent.

Now, let's fit a PGLS model using the correlation structure inferred with
`corPagel()`.

```{r}
#| warning: false

# Fit PGLS using a correlation structure inferred with `corPagel()`
spp <- rownames(barbet)
pgls_pagel <- fit_pgls(
    tree = barbet_tree, 
    data = log(barbet), 
    model_formula = "Lnalt ~ wing", 
    corStruct = corPagel(1, phy = barbet_tree, form = ~spp)
)

pgls_pagel

# Compare all models
rbind(
    model_comparison,
    data.frame(
        Model = "PGLS_Pagel",
        Slopes = pgls_pagel$coefficients[2],
        P_values = pgls_model$tTable[2, 4],
        row.names = NULL
    )
)
```

`corPagel()` estimates a $\lambda$ that is slightly larger than 1, which leads
to a model with a slightly smaller slope, but P-values remain the same.

---

2. If you multiply all the branches of your phylogenetic tree by 100, will 
your PGLS analysis change? Why or why not? Can you confirm this using R code?

```{r}
#| warning: false

# Create a new tree with branch lengths multiplied by 100
new_tree <- barbet_tree
new_tree$edge.length <- new_tree$edge.length * 100

# Fit PGLS
pgls_longbranches <- fit_pgls(
    tree = new_tree, 
    data = log(barbet), 
    model_formula = "Lnalt ~ wing", 
    corStruct = corBrownian(phy = barbet_tree, form = ~spp)
)

pgls_longbranches

# Are model coefficients and p-values the same?
identical(pgls_model$tTable, pgls_longbranches$tTable)
```

After multiplying branch lengths by 100, model coefficients and P-values
remain the same. Since the species relationships (i.e., tree topology)
were not affected and all branches were multiplied by 100, 
the variance-covariance matrix (i.e., correlation structure) remained the same,
which explains why nothing changed.

---

3. Use the data files from chapter 1 to run a phylogenetic ANCOVA for 
anoles testing for the effect of body size ("SVL") and ecomorphological state
("ecomorph") on forelimb length ("FLL") in anoles, using an 
Ornstein–Uhlenbeck model, as implemented in the ape function `corMartins()`, 
as your correlational structure for the residual error of the model. 
You can use the data files from chapter 1, but you will need to do some 
work to combine data across files.

```{r}
# Get phenotypic data
anolis <- merge(
    read.csv("http://www.phytools.org/Rbook/3/ecomorph.csv", row.names = 1),
    read.csv("http://www.phytools.org/Rbook/3/anole.data.csv", row.names = 1),
    by = "row.names"
) |>
    tibble::column_to_rownames("Row.names")

# Get tree
anolis_tree <- read.tree("http://www.phytools.org/Rbook/3/Anolis.tre")


# Check data for inconsistencies
check <- geiger::name.check(anolis_tree, anolis)

anolis_tree <- drop.tip(anolis_tree, check$tree_not_data)

# Fit PGLS - phylogenetic ANCOVA
spp <- rownames(anolis)
anolis_ancova <- fit_pgls(
    tree = anolis_tree,
    data = anolis,
    model_formula = "FLL ~ SVL + ecomorph",
    corStruct = corMartins(1, phy = anolis_tree, form = ~spp)
)

anova(anolis_ancova)
```

The phylogenetic ANCOVA shows that there is a significant effect of 
body size on forelimb length after controlling for the also significant 
effect of ecomorphological state.

## References {.unnumbered}
