
# Modeling continuous character evolution on a phylogeny

Loading required packages:

```{r load-pkgs}
#| message: false
#| warning: false

set.seed(123) # for reproducibility

library(phytools)
library(ape)
library(geiger)
library(dplyr)
```


## Practice problems

1. The data set used in this chapter also includes estimates 
of GC content (*"GC_Content_Percent"*). Fit the three models (BM, OU, and EB) 
for this character. Also test for phylogenetic signal in GC content. 
What do you conclude?

```{r}
# Get data
bacteria <- read.csv(
    "http://www.phytools.org/Rbook/4/bac_rates.csv", row.names = 1
)

# Get tree
bacteria_tree <- read.tree("http://www.phytools.org/Rbook/4/bac_rates.phy")

# Check tree and data for inconsistencies
check <- geiger::name.check(bacteria_tree, bacteria)
check

# Fit models
gc_content <- setNames(log(bacteria$GC_Content_Percent), rownames(bacteria))

## Brownian motion
model_bm <- fitContinuous(bacteria_tree, gc_content)

## Ornstein-Uhlenbeck
model_ou <- fitContinuous(
    bacteria_tree, gc_content, model = "OU", bounds = list(alpha = c(0, 10))
)

## Early-burst
model_eb <- fitContinuous(bacteria_tree, gc_content, model = "EB")

# Comparing the models
model_comp <- data.frame(
    Models = c("BM", "OU", "EB"),
    AIC = c(AIC(model_bm), AIC(model_ou), AIC(model_eb))
)
model_comp$AIC.w <- aic.w(model_comp$AIC)

model_comp
```

The Akaike Information Criteria (AIC) and AIC weights shows that the OU model
has more support.

Now, let's test for phylogenetic signal in GC content.

```{r}
# Test for phylogenetic signal
## Blomberg's K
signal_k <- phylosig(
    bacteria_tree, gc_content, method = "K", test = TRUE, nsim = 10000
)

## Pagel's lambda
signal_lambda <- phylosig(
    bacteria_tree, gc_content, method = "lambda", test = TRUE, nsim = 1000
)

# Summarizing model results
signal_summary <- data.frame(
    Method = c("Blomberg's K", "Pagel's lambda"),
    Stat = c(signal_k$K, signal_lambda$lambda),
    P_value = c(signal_k$P, signal_lambda$P)
)
signal_summary
```

Both statistics (Pagel's $\lambda$ and Blomberg's K) indicate that there is
stronger phylogenetic signal that what is expected by chance, which
means that the evolution of GC content is influenced by the shared
evolutionary history among species in the tree.


---

2. Reanalyze the data for mutation accumulation rate, but this time do not
log-transform the data. What happens? Can you explain the discrepancy?

```{r}
# Get mutation accumulation rate
mutation <- setNames(bacteria$Accumulation_Rate, rownames(bacteria))

# Fit model with log-transformed and non-log-transformed data
mutation_original <- fitContinuous(bacteria_tree, mutation)
mutation_log <- fitContinuous(bacteria_tree, log(mutation))

# Compare models
model_comp <- data.frame(
    Models = c("Log-transformed", "Original"),
    AIC = c(AIC(mutation_log), AIC(mutation_original))
)
model_comp$AIC.w <- aic.w(model_comp$AIC)

model_comp
```

When fitting a Brownian motion model on highly left-skewed data,
the parameter estimates change dramatically ($\sigma^2 = 0.00000$ and
$z_0 - 0.000004$). The reason for this discrepancy can be attributed to 
the highly left-skewed nature of the original data. Brownian motion assumes 
that the variable evolves continuously over time with a constant variance.
However, if the data is highly left-skewed, it indicates a lack of variation 
or change over time. This results in the model fitting a variance close to 
zero, indicating no significant evolutionary pattern.

---

3. You might wonder about the relationship between the models considered here 
and phylogenetic signal. Use a simulation study to find out! Focus for now 
on the OU model, which can be simulated using `fastBM()` in *phytools*. 
For simplicity, you can use the bacterial tree, `bacteria.tree`, 
that we used in this exercise. On that tree, simulate OU characters 
with $\sigma^2 = 1$, expected mean $\theta = 0$, and root state $\alpha = 0$. 
Vary $\alpha$ over a range of values from small $\alpha = 0.1$ to 
large $\alpha = 10$. For each simulation, determine the amount of 
phylogenetic signal, and do a significance test. See if you can detect the 
general pattern!

```{r}
# Get tree
bacteria_tree <- read.tree("http://www.phytools.org/Rbook/4/bac_rates.phy")

# Define simulation parameters
sigma2 <- 1
theta <- 0
alpha <- c(0.1, 0.5, 1, 2, 5, 10)
nsim <- 50

#' Simulate 50 OU characters with increasingly large alpha values and test
#' for phylogenetic signal in each simulation
sim <- Reduce(rbind, lapply(alpha, function(x) {
    
    # Simulate OU characters
    characters <- fastBM(
        bacteria_tree, alpha = x, theta = theta, sig2 = sigma2, nsim = nsim
    )
    
    # Test for phylogenetic signal in each character and get P-values
    sig <- Reduce(rbind, apply(characters, 2, function(y) {
        s <- phylosig(bacteria_tree, y, test = TRUE, nsim = 10000)
        results <- data.frame(
            K = s$K, 
            P = s$P,
            alpha = x
        )
        return(results)
    }))
    
    return(sig)
}))

# Calculate mean K and percentage of significant K values for each alpha
sim |>
    group_by(alpha) |>
    summarise(
        MeanK = mean(K),
        `% significant` = sum(P < 0.05)
    )
```

The simulations show that the larger the value of $\alpha$, the smaller
the K and the percentage of significant K values. Since larger values of
$\alpha$ indicate stronger selection pressures towards the optimum value,
the weaker phylogenetic signal is very reasonable, as stronger selection
leads to a more constrained evolution, reducing the influence of shared
evolutionary history on the trait values.

